Testing the social, cognitive, and motor foundations of Paleolithic skill reproduction

### 2021-08-21 manuscript/manuscript.Rmd
updated the text partially.

---
title: Testing the motor and cognitive foundations of Paleolithic social transmission
author: 
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
- Megan Beney Kilgore^[Department of Anthropology, Emory University, Atlanta, GA, USA; megan.elizabeth.beney@emory.edu]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Aditi Majoe^[Indepedent Researcher; am2752@cantab.ac.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
abstract: "Stone tools provide key evidence of human cognitive evolution but remain difficult to interpret. Toolmaking skill-learning in particular has been understudied even though: 1) the most salient cognitive demands of toolmaking should occur during learning, and 2) variation in learning aptitude would have provided the raw material for any past selection acting on tool making ability. However, we actually know very little about the cognitive prerequisites of learning under different information transmission conditions that may have prevailed during the Paleolithic. This paper presents results from a pilot experimental study to trial new experimental methods for studying the effect of learning conditions and individual differences on Oldowan flake-tool making skill acquisition. We trained 23 participants for 2 hours to make stone flakes under two different instructional conditions (observation only vs. direct active teaching) employing appropriate raw materials, practice time, and real human interaction. Participant performance was evaluated through analysis of the stone artifacts produced. Performance was compared both across experimental groups and with respect to individual participant differences in grip strength, motor accuracy, and cognitive function measured for the study. Our results show aptitude to be associated with fluid intelligence in a verbally instructed group and with a tendency to use social information in an observation-only group. These results have implications for debates surrounding the cumulative nature of human culture, the relative contributions of knowledge and know-how for stone tool making, and the role of evolved psychological mechanisms in “high fidelity” transmission of information, particularly through imitation and teaching.\\par \\textbf{Keywords:} Oldowan; Stone toolmaking; Social learning; Individual variation; Cognitive aptitudes; Motor skills"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

Stone tools have long been seen as a key source of evidence for understanding human behavioral and cognitive evolution [@darwin1871; @oakley1949; @washburn1960]. Pathbreaking attempts to infer specific cognitive capacities from this evidence largely focused on the basic requirements of tool production [@isaac1976; @wynn1979; @gowlett1984; @wynn2004]. More recently, increasing attention has been directed to the processes and demands of stone tool making skill acquisition [@roux1995; @stout2002; @stout2005; @geribàs2010; @nonaka2010; @stout2011; @putt2014; @hecht2015a; @duke2015; @morgan2015; @stout2015; @lombao2017; @putt2017; @cataldo2018; @putt2019; @pargeter2019; @pargeter2020]. This is motivated by the expectation that the most salient cognitive demands of tool making should occur during learning rather than routine expert performance [@stout2015] and by interest in the relevance of different social learning mechanisms such as imitation [@rein2014; @stout2019], emulation [@tehrani2008; @wilkins2018], and language [@ohnuma1997; @putt2014; @morgan2015; @lombao2017; @putt2017; @cataldo2018] to the reproduction of Paleolithic technologies.

Studies investigating these questions have used a range of different experimental designs (e.g., varying technological goals/instructions, training times, raw materials, live vs. recorded instruction, lithic/skill assessment metrics, pseudo-knapping tasks etc.) and reached disparate conclusions regarding the neurocognitive and social foundations of skill acquisition. It is plausible that these discordant results reflect actual diversity in how humans acquire and master stone tool making skills. However, this failure of results to generalize across artificial experimental manipulations [cf. @yarkoni2020] also raises doubts regarding the external validity [@eren2016] of conclusions with respect to real-world Paleolithic learning contexts. To address this, we conducted an exploratory study that draws on lessons from previous research in an attempt to balance the pragmatic and theoretical tradeoffs inherent in experimental studies of stone knapping skill acquisition [@pargeter2019; @stout2015].

Learning real-world skills like stone knapping is highly demanding of time and materials and difficult to control experimentally without sacrificing generalizability to real world conditions. Prior efforts have attempted to navigate these challenges by using various combinations of 1) inauthentic raw materials that are less expensive, easier to standardize, and/or easier to knap, 2) video-recorded instruction that is uniform across participants and less demanding of experimenter time, 3) short learning periods, 4) small sample sizes, and 5) single learning conditions. The difficulty of interpreting results from this growing literature led Stout and Khreisheh [-@stout2015: 870, emphasis original] to call for "studies with sufficient sample sizes to manipulate learning conditions (e.g. instruction, motivation) and assess individual variation (e.g. performance, psychometrics, neuroanatomy) that *also* have realistic learning periods." The current study attempts to strike a viable balance between these demands by investigating early-stage learning of a relatively simple technology (least effort, "Oldowan," flake production [@reti2016; @shea2016] under two instructional conditions while collecting data on individual differences in strength, coordination, cognition, social learning, self-control, and task engagement. Unlike any previous study, this allows us to address the likelihood that group effects of training conditions might be impacted by interactions with individual participant differences in aptitude, motivation, or learning style.

We focus on early stage learning because it has been found to be relatively rapid, variable across individuals, and predictive of later outcomes [@pargeter2019; @stout2015; @putt2019], and thus provides a reasonable expectation of generating meaningful data on skill and learning variation while minimizing training costs. Moreover, understanding the minimum training times necessary to detect changes in tool making skill will help archaeologists design more realistic and cost-effective experiments. To further manage costs, we limited our study to only two learning conditions (observation only vs. active teaching). This targets a key controversy in human evolution, namely the origins of teaching and language [@gärdenfors2017; @morgan2015], while avoiding highly artificial manipulations of dubious relevance to real-world Paleolithic learning. These choices allowed us to invest more in other aspects of research design that we identified as theoretically important, including measurement of individual differences in cognition and behavior, inclusion of an in-person, fully interactive teaching condition, and use of naturalistic raw materials. Sample size remained small in this internally funded exploratory study but could easily be scaled up at funding levels typical of pre- and post-doctoral research grants in archaeology.

@@

Individuals vary in aptitude and learning style for particular skills [@jonassen1993] but this has largely been ignored in studies of knapping skill acquisition, which have instead focused on group effects of different experimental conditions. There are good pragmatic reasons for this, as individual difference studies typically require larger sample sizes and additional data collection. However, overlooking these distinctions is not ideal since individual differences can provide valuable insight into the mechanisms, development, and evolution of cognition and behavior [@boogert2018]. In particular, patterns of association between cognitive traits and behavioral performance can be used to test hypotheses about the cognitive demands of learning particular skills and the likely targets of natural selection acting on aptitude. More prosaically, individual differences can introduce an unexamined and uncontrolled source of variation in group level results. This is especially true in the relatively small "samples of convenience" typical of experimental archaeology.

While testing hypotheses in evolutionary cognitive archaeology remains a considerable challenge [@wynn2017], investigation of individual variation in modern research participants represents one promising direction. For any particular behavior of archaeological interest, it is expected that standing variation in modern populations should remain relevant to normal variation in learning aptitude. The presence of trait variation without impact on learning aptitude would provide strong evidence against the plausibility of the proposed evolutionary relationship. An absence of variation (i.e., past fixation and rigorous developmental canalization) is not expected given the known variability of human brains and cognition [@sherwood2017; @barrett2020]. Any confirmatory findings of trait-aptitude correspondence would then have the testable implication that humans should be evolutionarily derived along the same dimension [e.g. @hecht2015b].

To date, a small number of "neuroarchaeological" studies have reported associations between individual knapping performance and brain structure or physiological responses. Hecht et al. [-@hecht2015a] reported training-related changes in white matter integrity (fractional anisotropy [FA]) that correlated with individual differences in practice time and striking accuracy change. The regional patterning of FA changes also varied across individuals, with only those individuals who displayed early increases in FA under the right ventral precentral gyrus (premotor cortex involved in movement planning and guidance) showing striking accuracy improvement over the training period. Putt et al. [-@putt2019] similarly found that the proportion of flakes to shatter produced by individuals during handaxe making correlated with dorsal precentral gyrus (motor cortex) activation. Pargeter et al. [-@pargeter2020] used a flake prediction paradigm [modeled after @nonaka2010] to confirm that striking force and accuracy are important determinants of handaxe-making success. These findings all point to the central role of perceptual-motor systems [@stout2007] and coordination [@roux1995] in knapping skill acquisition. In addition, Putt et al. [-@putt2019] also found successful flake production to be associated with prefrontal (working memory/cognitive control) activation and Stout et al. [-@stoutCognitiveDemandsLower2015] found that prefrontal activation correlated with success at a strategic judgement (platform selection) task which in turn was predictive of success at out-of-scanner handaxe production. Such investigations are thus starting to chart out the more specific contributions of different neural systems to particular aspects of knapping skill acquisition. To date, however, the cognitive/functional interpretation of systems identified in this manner has largely relied on informal reverse inference (reasoning backward from observed activations to inferred mental processes) from published studies of other tasks that activated the same regions, an approach which is widely regarded as problematic [@poldrack2011].

Here we take a more direct, psychometric approach to measuring individual differences in perceptual-motor coordination and cognition. Psychometric instruments (e.g., tasks, questionnaires) are designed to assess variation in cognitive traits and states, such as fluid intelligence, working memory, attention, motivation, and personality, that have been of theoretical interest to cognitive archaeologists [e.g., @wynn2016]. It is thus surprising that they have been almost entirely neglected in experimental studies of knapping skill. In the only published example we are aware of, Pargeter et al. [-@pargeter2019] reported significant effects of variation in planning and problem solving (Tower of London test [@shallice1982]) and cognitive set shifting (Wisconsin Card Sort test [@grant1948]) on early stage handaxe learning. Of course, cognition is not the only thing that can affect knapping performance. Flake prediction experiments highlight the importance of regulating movement speed/accuracy trade-offs [@nonaka2010; @pargeter2020] and studies of muscle recruitment [@marzke1998] and manual pressure [@williams-hatala2018; @key2018] during knapping highlight basic strength requirements. Along these lines, Key and Lycett [-@key2019] found that individual differences in hand size, shape, and especially grip strength were better predictors of force loading during stone tool use than were attributes of the tools themselves. However, we are unaware of any such studies of biometric influences on variation in knapping success. Finally, the time and effort demands of knapping skill acquisition suggest that differences in personality (e.g., self-control and "grit" [@pargeter2019], motivation [@stout2002], and social vs. individual learning strategies [@miu2020] might also affect learning outcomes. We are again unaware of any previous studies that have assessed such effects. In this study, we assessed all participants with a battery of tests including grip strength, movement speed/accuracy, spatial working memory, fluid intelligence, self-control, tendency to use social information, and motivation/engagement with the tool making task. We were particularly interested in the possibility that these variables might not only impact learning generally, but might also have different effects under different learning conditions.

@@

Possible links between tool making and language have been a subject of speculation for nearly 150 years [@engles2003[1873]], if not longer [@hewes1993], although compelling empirical tests have remained elusive. Over 25 years ago, Toth and Schick [-@toth1993] suggested that experiments teaching modern participants to make stone tools in verbal and non-verbal conditions could test the importance of language in the social reproduction of Paleolithic technologies. Ohnuma et al. [-@ohnuma1997] were the first to implement this suggestion in a study of Levallois flake production, followed by more recent studies of handaxe making [@putt2014; @putt2017] and simple flake production [@morgan2015; @cataldo2018; @lombao2017]. This reflects recent interest in the hypothesis that language might be an adaptation for teaching [e.g., @laland2017; @stout2012]. Teaching and learning demands of Paleolithic tool making would thus provide evidence of selective contexts favoring language evolution [@stout2010; @morgan2015; @montagu1976].

Toth and Schick [-@toth1993] were, however, careful to point out that extinct hominid learning strategies and capacities might differ from modern experimental participants. Even leaving aside potential species differences in social learning [cf. @morgan2015; @stout2019], reliance on explicit verbal instruction varies widely across modern human societies [e.g., @boyette2017]. The WEIRD (Western, educated, industrialized, rich, democratic [@henrich2010]) teachers and learners typical of knapping experiments arguably represent an extreme bias toward such instruction. Simply instructing such participants not to speak during an experiment (or to demonstrate but not gesture, etc. [@morgan2015]) is likely to underestimate the efficacy of non-verbal teaching and learning in cultural contexts where it is more common, let alone in a hypothetical pre-linguistic hominid species.

Such concerns are exacerbated in experiments using pre-recorded instructional videos or extremely short training periods. Video does not allow the interactive teaching that is favored even in formal academic knapping classes [e.g., @shea2015] and is almost certainly typical of traditional learning contexts [e.g., @stout2002]. It is not known how video presentation affects the efficacy of teaching generally, or the relative effectiveness of different forms of instruction. Going further, some experiments have manipulated the presence/absence of verbal instruction by presenting the same video with and without sound [@putt2017] or the sound track without the video [@cataldo2018]. While this provides experimental control, it does not allow the instructor to adjust their multi-modal [@levinson2014] communication strategies as they would naturally do, for example through pointing and pantomime. To simply remove a communication channel without allowing any such adaptation is highly artificial and risks generating results that cannot be generalized beyond the specific context of the experiment [@yarkoni2020]. Similarly, unnaturally short training periods (e.g., 5-15 minutes [@morgan2015; @lombao2017]) might misrepresent the relative efficacy of different teaching strategies under more realistic conditions [@whiten2015; @stout2015]. Even the longest training times to date [@stout2015; @pargeter2019] have not produced knapping skills comparable to relevant archaeological examples, and were achieved by limiting sample size and using only one teaching condition.

For these reasons, we sought to explore a middle path between experimental expedience and realism by limiting our experiment to two relatively naturalistic learning conditions and a moderate learning period of two hours. As in previous experiments [@stout2011; @hecht2015a; @pargeter2019] the first condition was unrestricted, interactive instruction in small groups, essentially reproducing the "natural" teaching/learning context familiar [cf. @shea2015] to our WEIRD instructor and student participants. The second condition allowed observation only, with the experimenter visible making flakes but not interacting in any way with learners. This absence of teaching is again a familiar social context for our participants and did not require any novel behaviors from the instructor. It matches the "imitation/emulation" condition of Morgan et al. [-@morgan2015] although we make no assumptions regarding learning mechanisms. We did not include a "reverse engineering" or "end-state emulation" condition in which only finished products were visible. This has been advocated as an important baseline or control condition [@whiten2015] to distinguish observational from individual learning, but is not likely to model any typical Paleolithic learning context nor to stand as an adequate proxy for the cognition of hominid species with different social learning capacities. There is no reason to assume neurocognitive and behavioral processes of reverse-engineering problem solving in modern humans [e.g., @allen2020] approximate the social learning processes of hominids with more ape-like action observation/imitation capacities [@hecht2013a; @hecht2013b; @stout2019].

We selected a two-hour learning period for both pragmatic and theoretical reasons. Pargeter et al. [@pargeter2019] found that even \~90 hours of fully interactive instruction and practice was insufficient to achieve handaxe-making skills comparable to the later Acheulean site of Boxgrove [@garcía-medrano2019; @stout2014], and estimated actual time to mastery as ranging from 121 to 441 hours for different participants. However, they observed the greatest, fastest, and most individually variable skill increases during the first 20 hours of practice. In addition, initial performance was moderately correlated with later achievement. This suggests that studying early-stage learning may be a pragmatic alternative, especially for research investigating individual differences in aptitude. Studies of simple flake production similarly document large initial variation [@stout2015] and rapid early progress [@stout2015; @putt2019; @stoutKnappingSkillEarliest2006]. We designed the current study to test the utility of studying learning and variation during the first two hours of simple flaking instruction/practice, in hopes of finding a viable compromise between experimental realism and cost.

@@

Lithic raw materials vary in size, shape, and fracture mechanical properties that affect the difficulty of achieving different knapping goals [@eren2014]. Unfortunately, it can be difficult and/or expensive to procure authentic raw materials. Experimental studies of knapping skill have often used proxy materials such as flint [@morgan2015; @nonaka2010; @cataldo2018], limestone [@stoutKnappingSkillEarliest2006], porcelain [@khreisheh2013], or heat-treated chert [@putt2014; @putt2017; @putt2019]to model Oldowan and early Acheulean technologies executed in other materials. As well as being more readily available, these proxies are generally easier to knap. This has the benefit of reducing required practice time, but it is unclear how it might affect learning demands more generally or the efficacy of different learning conditions/strategies specifically.

>>>>

---
title: Testing the social, cognitive, and motor foundations of Paleolithic skill reproduction
author: 
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Megan Beney Kilgore^[Department of Anthropology, Emory University, Atlanta, GA, USA; megan.elizabeth.beney@emory.edu]
- Aditi Majoe^[Indepedent Researcher; am2752@cantab.ac.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
abstract: "Stone tools provide key evidence of human cognitive evolution but remain challenging to interpret. Tool-making skill-learning has been understudied even though: 1) the most salient cognitive demands of tool-making should occur during learning, and 2) variation in learning aptitude would have provided the raw material for any past selection acting on tool-making ability. However, we know very little about the cognitive prerequisites of learning under different information transmission conditions that may have prevailed during the Paleolithic. This paper presents results from an exploratory study to trial new experimental methods for studying the effect of learning conditions and individual differences on Oldowan flake-tool-making skill acquisition. We trained 23 participants for 2 hours to make stone flakes under two different instructional conditions (observation only vs. direct active teaching). We employed appropriate raw materials, a moderate practice time duration, and in-person, fully interactive instruction. Participant performance was evaluated through an analysis of the stone artifacts produced. We compared performance across experimental groups with respect to individual participant differences in grip strength, motor accuracy, and cognitive function measured for the study. Our results show that 2 hours are insufficient to document learning-related performance change. However, teaching reduces variability in knapping rate, methods, and outcomes during early-stage learning, thus increasing the reliability of skill reproduction. Instruction also altered knapping quality vs. quantity trade-offs in the two groups and dramatically changed the effects of individual differences in strength, visuospatial working memory, and social learning tendencies on knapping outcomes. Our results provide further support for the hypothetical co-evolution of teaching, language, and tool-making, suggest that the presence/absence of instruction can fundamentally alter learning-related selection pressures on individuals, and offer lessons for the design of future experiments.\\par \\textbf{Keywords:} Oldowan; Stone tool-making; Social learning; Individual variation; Cognitive aptitudes; Motor skills"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

Stone tools have long been seen as a key source of evidence for understanding human behavioral and cognitive evolution [@darwin1871; @oakley1949; @washburn1960]. Pathbreaking attempts to infer specific cognitive capacities from this evidence largely focused on the basic requirements of tool production [@isaac1976; @wynn1979; @gowlett1984; @wynn2004]. More recently, researchers have directed increasing attention to the processes and demands of stone tool-making skill acquisition [@roux1995; @stout2002; @stout2005; @geribàs2010; @nonaka2010; @stout2011; @putt2014; @hecht2015a; @duke2015; @morgan2015; @stout2015; @lombao2017; @putt2017; @cataldo2018; @putt2019; @pargeter2019; @pargeter2020; @shipton2018]. This increased attention is motivated by the expectation that tool-making's most salient cognitive demands should occur during learning rather than routine expert performance [@stout2015]. Researchers are also increasingly focused on the relevance of different social learning mechanisms such as imitation [@rein2014; @stout2019], emulation [@tehrani2008; @wilkins2018], and language [@ohnuma1997; @putt2014; @morgan2015; @lombao2017; @putt2017; @cataldo2018] to the reproduction of Paleolithic technologies.

Studies investigating these questions have used a range of different experimental designs (e.g., varying technological goals/instructions, training times, raw materials, live vs. recorded instruction, lithic/skill assessment metrics, pseudo-knapping tasks etc.) and reached disparate conclusions regarding skill acquisition’s neurocognitive and social foundations. It is plausible that these discordant results reflect actual diversity in how humans acquire and master stone tool-making skills. However, this failure of results to generalize across artificial experimental manipulations [cf. @yarkoni2020] also raises doubts regarding the external validity [@eren2016] of conclusions concerning real-world Paleolithic learning contexts. To address this, we conducted an exploratory study that draws on lessons from previous research to balance the pragmatic and theoretical trade-offs inherent in experimental studies of stone knapping skill acquisition [@pargeter2019; @stout2015].

Compared to artificial experimental tasks, learning real-world skills like stone knapping is highly demanding of time and materials and difficult to control experimentally without sacrificing generalizability to real world conditions. Prior efforts have attempted to navigate these challenges by using various combinations of 1) inauthentic raw materials that are less expensive, easier to standardize, and/or easier to knap, 2) video-recorded instruction that is uniform across participants and less demanding of experimenter time, 3) short learning periods, 4) small sample sizes, and 5) single learning conditions. The difficulty of interpreting results from this growing literature led Stout and Khreisheh [-@stout2015: 870, emphasis original] to call for "studies with sufficient sample sizes to manipulate learning conditions (e.g. instruction, motivation) and assess individual variation (e.g. performance, psychometrics, neuroanatomy) that *also* have realistic learning periods." The current study attempts to strike a viable balance between these demands by investigating early-stage learning of a relatively simple technology (least effort, "Oldowan," flake production [@reti2016; @shea2016]) under two instructional conditions while collecting data on individual differences in strength, coordination, cognition, and social learning. Unlike any previous study, this allows us to address the likelihood that interactions with individual participant differences in aptitude or learning style might impact group effects of training conditions.

We focus on early-stage learning because it is relatively rapid, variable across individuals, and predictive of later outcomes [@pargeter2019; @stout2015; @putt2019]. The study thus provides a reasonable expectation of generating meaningful data on skill and learning variation while minimizing training costs. Moreover, understanding the minimum training times necessary to detect changes in tool making skill will help archaeologists design more realistic and cost-effective experiments. We limited our study to only two learning conditions (observation only vs. active teaching) to further manage costs. This study design targets a key controversy in human evolution, namely the origins of teaching and language [@gärdenfors2017; @morgan2015] while avoiding artificial manipulations of dubious relevance to real-world Paleolithic learning [e.g., @cataldo2018]. These choices allowed us to invest more in other aspects of research design that we identified as theoretically necessary, including measuring individual differences in cognition and behavior, including an in-person, fully interactive teaching condition, and using naturalistic raw materials. The sample size remained small in this internally funded exploratory study, but we could easily scale it up at funding levels typical of pre- and post-doctoral research grants in archaeology.

@@

Individuals vary in aptitude and learning style for skills [@jonassen1993], but research on knapping skill acquisition has largely ignored this. These studies have instead focused on group effects of different experimental conditions. There are good pragmatic reasons for this, as individual difference studies typically require larger sample sizes and other data collection. However, overlooking these distinctions is not ideal since individual differences can provide valuable insight into cognition and behavior's mechanisms, development, and evolution [@boogert2018]. Researchers can use patterns of association between cognitive traits and behavioral performance to test hypotheses about the cognitive demands of learning particular skills and the likely targets of natural selection acting on aptitude. More prosaically, individual differences can introduce an unexamined and uncontrolled source of variation in group-level results. This effect is especially true in the relatively small samples of convenience typical of experimental archaeology.

While testing hypotheses in evolutionary cognitive archaeology remains a considerable challenge [@wynn2017], investigating individual variation in modern research participants represents one promising direction. For any behavior of archaeological interest, we could expect that standing variation in modern populations should remain relevant to normal variation in learning aptitude. The presence of trait variation without impact on learning aptitude would provide strong evidence against the plausibility of the proposed evolutionary relationship. We would not expect an an absence of variation (i.e., past fixation and rigorous developmental canalization) given the known variability of human brains and cognition [@sherwood2017; @barrett2020]. Any confirmatory findings of trait-aptitude correspondence would have the testable implication that humans should be evolutionarily derived along the same dimension [@hecht2015b].

To date, a small number of "neuroarchaeological" studies have reported associations between individual knapping performance and brain structure or physiological responses. Hecht et al. [-@hecht2015a] reported training-related changes in white matter structure correlated with individual differences in practice time and striking accuracy change. The regional patterning of white matter changes also varied across individuals. Only those who displayed early increases under the right ventral precentral gyrus (premotor cortex involved in movement planning and guidance) showed striking accuracy improvement over the training period. Putt et al. [-@putt2019] similarly found that the proportion of flakes-to-shatter produced by individuals during handaxe-making correlated with dorsal precentral gyrus (motor cortex) activation. Pargeter et al. [-@pargeter2020] used a flake prediction paradigm [modeled after @nonaka2010] to confirm that striking force and accuracy are important determinants of handaxe-making success. These findings point to the central role of perceptual-motor systems [@stout2007] and coordination [@roux1995] in knapping skill acquisition. In addition, Putt et al. [-@putt2019] found successful flake production to be associated with prefrontal (working memory/cognitive control) activation, and Stout et al. [-@stoutCognitiveDemandsLower2015] found that prefrontal activation correlated with success at a strategic judgement (platform selection) task which in turn was predictive of success at out-of-scanner handaxe production.

Such investigations are thus starting to chart the contributions of different neural systems to aspects of knapping skill acquisition, including motor and cognitive control aspects. To date, however, the cognitive/functional interpretation of systems identified in this manner has relied mainly on informal reverse inference (reasoning backward from observed activations to inferred mental processes) from published studies of other tasks that activated the same regions. Researchers widely regard this approach as problematic [@poldrack2011]. Here we take a more direct, psychometric approach to measuring individual differences in perceptual-motor coordination and cognition.

Studies implement psychometric instruments (e.g., tasks, questionnaires) to assess variation in cognitive traits and states, such as fluid intelligence, working memory, attention, motivation, and personality. These measures have been of theoretical interest to the broader field of cognitive archaeology [e.g., @wynn2016]. It is thus surprising that experimental studies have almost entirely neglected them. In the only published example we are aware of, Pargeter et al. [-@pargeter2019] reported significant effects of variation in planning and problem solving (Tower of London test [@shallice1982]) and cognitive set shifting (Wisconsin Card Sort test [@grant1948]) on early stage handaxe learning. Of course, cognition is not the only thing that can affect knapping performance. Flake prediction experiments highlight the importance of regulating movement speed/accuracy trade-offs [@nonaka2010; @pargeter2020] and studies of muscle recruitment [@marzke1998] and manual pressure [@williams-hatala2018; @key2018] during knapping highlight basic strength requirements. Along these lines, Key and Lycett [-@key2019] found that individual differences in hand size, shape, and especially grip strength were better predictors of force loading during stone tool use than were attributes of the tools themselves. However, we are unaware of any such studies of biometric influences on variation in knapping success. Finally, individual, and contextual differences in adopting social vs. individual learning strategies [@miu2020] might also affect learning outcomes. We are again unaware of any previous studies that have assessed such effects. This study evaluated all participants with a battery of tests, including grip strength, movement speed/accuracy, spatial working memory, fluid intelligence, and tendency to use social information. We were particularly interested in the possibility that these variables might not only impact learning generally but might also have different effects under different learning conditions.

@@

Possible links between tool making and language have been a subject of speculation for nearly 150 years [@engles2003[1873]], if not longer [@hewes1993], although compelling empirical tests have remained elusive. Over 25 years ago, Toth and Schick [-@toth1993] suggested that experiments teaching modern participants to make stone tools in verbal and non-verbal conditions could test the importance of language in the social reproduction of Paleolithic technologies. Ohnuma et al. [-@ohnuma1997] were the first to implement this suggestion in a study of Levallois flake production, followed by more recent studies of handaxe making [@putt2014; @putt2017] and simple flake production [@morgan2015; @cataldo2018; @lombao2017]. This study design trajectory reflects a recent interest in the hypothesis that language might be an adaptation for teaching [e.g., @laland2017; @stout2012]. Teaching and learning demands of Paleolithic tool-making would thus provide evidence of selective contexts favoring language evolution [@stout2010; @morgan2015; @montagu1976].

However, Toth and Schick [-@toth1993] pointed out that extinct hominid learning strategies and capacities might differ from modern experimental participants. Even leaving aside potential species differences in social learning [cf. @morgan2015; @stout2019], reliance on explicit verbal instruction varies widely across modern human societies [e.g., @boyette2017]. The WEIRD (Western, educated, industrialized, rich, democratic [@henrich2010]) teachers and learners typical of knapping experiments arguably represent an extreme bias toward such instruction. Simply instructing such participants not to speak during an experiment (or to demonstrate but not gesture, etc. [@morgan2015]) is likely to underestimate the efficacy of non-verbal teaching and learning in cultural contexts where it is more common, let alone in a hypothetical pre-linguistic hominid species [cf. @fuentes2015].

Such concerns are exacerbated in experiments using pre-recorded instructional videos or brief training periods. Video does not allow the interactive teaching favored even in formal academic knapping classes [e.g., @shea2015] and is almost undoubtedly typical of traditional learning contexts [e.g., @stout2002]. Researchers do not know how video presentation generally affects the efficacy of teaching stone knapping or the relative effectiveness of different forms of instruction. Moreover, some experiments have manipulated the presence/absence of verbal instruction by presenting the same video with and without sound [@putt2017] or the sound track without the video [@cataldo2018]. While this provides experimental control, it does not allow the instructor to adjust their multimodal [@levinson2014] communication strategies as they would naturally do, for example, through pointing and pantomime. Simply removing a communication channel without allowing such adaptation is highly artificial and risks generating results that cannot be generalized beyond the specific context of the experiment [@yarkoni2020]. Similarly, unnaturally short training periods (e.g., 5-15 minutes [@morgan2015; @lombao2017]) might misrepresent the relative efficacy of different teaching strategies under more realistic conditions [@whiten2015; @stout2015]. Even the most extended training times [@stout2015; @pargeter2019] have not produced knapping skills comparable to relevant archaeological examples and were achieved by limiting sample size and using only one teaching condition.

For these reasons, we sought to explore a middle path between experimental expedience and realism by limiting our experiment to two relatively naturalistic learning conditions and a moderate learning period of two hours. As in previous experiments [@stout2011; @hecht2015a; @pargeter2019], the first condition was unrestricted, interactive instruction in small groups, essentially reproducing the "natural" teaching/learning context familiar [cf. @shea2015] to our WEIRD instructor and student participants. The second condition allowed observation only, with the experimenter visible making flakes but not providing instruction or interacting with learners in any way during this activity. This absence of teaching is again a familiar social context for our participants and did not require any novel behaviors from the instructor. Although we make no assumptions regarding learning mechanisms, it matches the "imitation/emulation" condition of Morgan et al. [-@morgan2015]. We did not include a "reverse engineering" or "end-state emulation" condition where only finished products were visible. Researchers have advocated these states as an important baseline or control condition [@whiten2015] to distinguish observational from individual learning. However, they are not likely to model any typical Paleolithic learning context nor to stand as an adequate proxy for the cognition of hominid species with different social learning capacities. There is no reason to assume neurocognitive and behavioral processes of reverse-engineering problem solving in modern humans [e.g., @allen2020] approximate the social learning processes of hominids with more ape-like action observation/imitation capacities [@hecht2013a; @hecht2013b; @stout2019].

We selected a two-hour learning period for both pragmatic and theoretical reasons. Pargeter et al. [@pargeter2019] found that even \~90 hours of fully interactive instruction and practice was insufficient to achieve handaxe-making skills comparable to the later Acheulean site of Boxgrove [@garcía-medrano2019; @stout2014], and estimated actual time to mastery as ranging from 121 to 441 hours for different participants. However, they observed the greatest, fastest, and most individually variable skill increases during the first 20 hours of practice. In addition, initial performance moderately correlated with later achievement. This result suggests that studying early-stage learning may be a practical alternative, especially for research investigating individual differences in aptitude. Studies of simple flake production similarly document considerable initial variation [@stout2015] and rapid early progress [@stout2015; @putt2019; @stoutKnappingSkillEarliest2006]. We designed the current study to test the utility of studying learning and variation during the first two hours of simple flaking instruction/practice in hopes of finding a viable compromise between experimental realism and cost.

@@

Lithic raw materials vary in size, shape, and fracture mechanical properties that affect the difficulty of achieving different knapping goals [@eren2014]. Unfortunately, it can be difficult and/or expensive to procure authentic raw materials. Experimental studies of knapping skill have often used proxy materials such as flint [@morgan2015; @nonaka2010; @cataldo2018], limestone [@stoutKnappingSkillEarliest2006], porcelain [@khreisheh2013], or heat-treated chert [@putt2014; @putt2017; @putt2019] to model Oldowan and early Acheulean technologies executed in other materials. As well as being more readily available, these proxies are generally easier to knap. This study design has the benefit of reducing required practice time. Still, it is unclear how it might affect learning demands more generally or the efficacy of different learning conditions/strategies precisely.

Some designs have attempted to closely match experimental and archaeological raw material types [@stout2011; @duke2015; @pargeter2019]. However, raw materials vary across individual clasts within as well as between types. This choice has led to an interest in standardizing experimental core morphology [@nonaka2010] and composition, even if this means using artificial materials such as porcelain [@khreisheh2013], brick [@geribàs2010; @lombao2017], or foam blocks [@schillinger2014]. Such manipulations enhance experimental control and internal validity [@eren2016] at the expense of external generalizability to actual archaeological conditions. Specifically, they allow more robust results from smaller samples but eliminate a core element of real-world knapping skill: the ability to produce consistent results from variable materials [@pelegrin1990; @stout2013]. For example, Pargeter et al. [-@pargeter2020] found that predicting specific flaking outcomes on actual handaxe preforms was both more complex and less technologically important than expected from previous work with standardized, frustum-shaped cores [@nonaka2010]. The alternative to control is to incorporate raw material size, shape, and composition as experimental variables [e.g., @stout2019]. This choice allows consideration of raw material selection and response to variation as aspects of skill but correspondingly increases the sample sizes required to identify patterning. In considering these issues, we again chose to explore a middle path between pragmatism and realism by employing commercially purchased basalt like that known from East African Oldowan sites. We allowed clast size and shape to vary within set limits and selected the particular clasts provided to each participant to approximate the same distribution.
